---
title: "Privacy utility trade off"
author: "Finta_Francis_R_project"
date: "2026-08-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

library(synthpop)

library(partykit)

library(pROC)

library(ggplot2)

## Load the dataset 

library(readr)
steam <- read_csv("~/Downloads/Blinded_data_sets/dataset30_ 9779 .csv")

## Ctree synthesis
set.seed(123)

n <- nrow(steam)

train_id <- sample(
  1:n,
  size = round(0.8 * n)
)

train_data <- steam[train_id, ]
test_data  <- steam[-train_id, ]

## Now create a ctree syn dataset

set.seed(123)

ctree_data <- syn(
  train_data,
  method = "ctree",
  m = 1
)$syn

## Now you have members which are train dataset 
## Non members - test dataset 
## synthetic dataset - ctree_data 

## Can I distinguish train_data from test_data using the synthetic dataset?
## This is the privacy question?

colnames(steam)
## So we got variables , we will assign these to a variable var 

vars <- c("Y", "x1", "x2", "x3", "x4", "x5", "x6", "W") ## ptid is removed 

## Now we will assign members, nonmembers, synthetic 

members <- train_data[, vars]

nonmembers <- test_data[, vars]

synthetic <- ctree_data[, vars]

head(train_data$ptid)
head(ctree_data$ptid)
## Now we will check the structure of the dataset
str(steam)
```
## Now we will calculate the distance between the variables 

```{r privacy}
library(cluster)

members_dist <- members
nonmembers_dist <- nonmembers
synthetic_dist <- synthetic

binary_vars <- c("Y", "x5", "W")

members_dist[binary_vars] <- lapply(
  members_dist[binary_vars],
  factor
)

nonmembers_dist[binary_vars] <- lapply(
  nonmembers_dist[binary_vars],
  factor
)

synthetic_dist[binary_vars] <- lapply(
  synthetic_dist[binary_vars],
  factor
)


## check again
str(members_dist) ## we have only numeric and factors here 

member_dist <- daisy(
  members_dist,
  metric = "gower"
)

nonmember_dist <- daisy(
  nonmembers_dist,
  metric = "gower"
)
## Now we will bind them

all_data <- rbind(members, nonmembers, synthetic)


gower_dist <- daisy(
  all_data,
  metric = "gower"
)

gower_matrix <- as.matrix(gower_dist)

n_members <- nrow(members)
n_nonmembers <- nrow(nonmembers)
n_synthetic <- nrow(synthetic)

member_rows <- 1:n_members

nonmember_rows <- 
  (n_members + 1):(n_members + n_nonmembers)

synthetic_rows <- 
  (n_members + n_nonmembers + 1):
  (n_members + n_nonmembers + n_synthetic)

## Finding closest record 

member_min_dist <- apply(
  gower_matrix[
    member_rows,
    synthetic_rows,
    drop = FALSE
  ],
  1,
  min
)

nonmember_min_dist <- apply(
  gower_matrix[
    nonmember_rows,
    synthetic_rows,
    drop = FALSE
  ],
  1,
  min
)


member_min_dist
nonmember_min_dist

## Ploting 

privacy_scores <- data.frame(
  distance = c(
    member_min_dist,
    nonmember_min_dist
  ),
  membership = c(
    rep("Member", length(member_min_dist)),
    rep("Non-member", length(nonmember_min_dist))
  )
)

ggplot(
  privacy_scores,
  aes(
    x = distance,
    fill = membership
  )
) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Distance to Nearest Synthetic Record",
    x = "Minimum Gower distance",
    y = "Density",
    fill = "Group"
  )
```
## plot suggests that this strategy could work to some extent, because members tend to have smaller distances.
## But because the curves overlap heavily, the attacker will make many mistakes.

## MIA-AUC 

```{r MIA}
attack_data <- data.frame(
  membership = c(
    rep(1, length(member_min_dist)),
    rep(0, length(nonmember_min_dist))
  ),
  score = c(
    -member_min_dist,
    -nonmember_min_dist
  )
)

library(pROC)

mia_roc <- roc(
  response = attack_data$membership,
  predictor = attack_data$score
)

auc(mia_roc)

## Plot 

plot(
  mia_roc,
  main = "Membership Inference Attack ROC Curve",
  xlab = "False Positive Rate (1 - Specificity)",
  ylab = "True Positive Rate (Sensitivity)",
  legacy.axes = TRUE
)

```
## Attacker advantage 

```{r AA}
best <- coords(
  mia_roc,
  x = "best",
  best.method = "youden",
  ret = c(
    "threshold",
    "sensitivity",
    "specificity"
  )
)

best

## calculate TPR, FPR, AA 

TPR <- best["sensitivity"]

FPR <- 1 - best["specificity"]

attack_advantage <- TPR - FPR

attack_advantage
```
## Calculating DCR
```{r DCR}
synthetic_min_dist <- apply(
  gower_matrix[
    synthetic_rows,
    member_rows,
    drop = FALSE
  ],
  1,
  min
)

mean(synthetic_min_dist)

## Higher DCR - better privacy

## DCR tries to find how close is synthetic data to the training dataset

```
## Calculate the NNDR
```{r NNDR}
synthetic_to_members <- gower_matrix[
  synthetic_rows,
  member_rows,
  drop = FALSE
] ## assign the matrix 

## Choose the closest two training records

synthetic_nndr <- apply(
  synthetic_to_members,
  1,
  function(x) {
    two_closest <- sort(x)[1:2]
    two_closest[1] / two_closest[2]
  }
)

synthetic_nndr

mean(synthetic_nndr)
```
## Since d 1 ≤d 2 , NNDR normally lies between 0 and 1.

## NNDR close to 0 → nearest training record is much closer than the second-nearest → potentially more suspicious.
## NNDR close to 1 → nearest and second-nearest records are similarly distant → less evidence that one particular training record is unusually close.

## Therefore, 0.8011 is relatively high, meaning that, on average, the closest and second-closest training records are not extremely different in distance.

```{r nndr2}
quantile(
  synthetic_nndr,
  probs = c(0, 0.01, 0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99, 1),
  na.rm = TRUE
) ## to check what happens in the tail
```
## IMS
```{r ims}
n_identical <- sum(synthetic_min_dist == 0)

n_identical
```
## none of your synthetic records is exactly identical to a record in the training/member dataset.

## Now let us compare 

```{r compare}
# ============================================
# SYNTHETIC -> NON-MEMBER DISTANCES
# ============================================

synthetic_to_nonmembers <- gower_matrix[
  synthetic_rows,
  nonmember_rows,
  drop = FALSE
]


# ============================================
# DCR
# ============================================

synthetic_min_dist_nonmember <- apply(
  synthetic_to_nonmembers,
  1,
  min
)

mean_dcr_nonmember <- mean(
  synthetic_min_dist_nonmember
)

dcr_5_train <- quantile(
  synthetic_min_dist,
  0.05,
  na.rm = TRUE
)

dcr_5_nonmember <- quantile(
  synthetic_min_dist_nonmember,
  0.05,
  na.rm = TRUE
)


# ============================================
# NNDR
# ============================================

synthetic_nndr_nonmember <- apply(
  synthetic_to_nonmembers,
  1,
  function(x) {

    two_closest <- sort(x)[1:2]

    if (two_closest[2] == 0) {
      return(NA_real_)
    }

    two_closest[1] / two_closest[2]
  }
)

mean_nndr_nonmember <- mean(
  synthetic_nndr_nonmember,
  na.rm = TRUE
)

nndr_5_train <- quantile(
  synthetic_nndr,
  0.05,
  na.rm = TRUE
)

nndr_5_nonmember <- quantile(
  synthetic_nndr_nonmember,
  0.05,
  na.rm = TRUE
)


# ============================================
# IMS
# ============================================

tolerance <- 1e-10

n_identical_nonmember <- sum(
  synthetic_min_dist_nonmember <= tolerance
)


# ============================================
# PRINT RESULTS
# ============================================

cat("DCR - mean training:", mean(synthetic_min_dist), "\n")
cat("DCR - mean non-member:", mean_dcr_nonmember, "\n\n")

cat("DCR - 5th percentile training:", dcr_5_train, "\n")
cat("DCR - 5th percentile non-member:", dcr_5_nonmember, "\n\n")

cat("NNDR - mean training:", mean(synthetic_nndr, na.rm = TRUE), "\n")
cat("NNDR - mean non-member:", mean_nndr_nonmember, "\n\n")

cat("NNDR - 5th percentile training:", nndr_5_train, "\n")
cat("NNDR - 5th percentile non-member:", nndr_5_nonmember, "\n\n")

cat("IMS - training:", n_identical, "\n")
cat("IMS - non-member:", n_identical_nonmember, "\n")
```
## The synthetic data has a stronger proximity to the training data than to the holdout data according to the DCR and NNDR comparisons.

## That can be evidence of potential privacy leakage/memorization, but it is not proof of an actual privacy breach.