---
title: "STEAM_project"
author: "Finta_Francis_R_project"
date: "2026-07-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

##Loading the dataset
library(readr)
steam<- read_csv("Downloads/Blinded_data_sets/dataset30_ 9779 .csv")

view(steam)

## Now let us explore the variables, look into what we have in out dataset

summary(steam)
str(steam)

## This gave us the information that whether the variables are binary or continuous, ot characters etc

sapply(steam,class)
## Now we know all are numeric 

ncol(steam)
## 9 columns with one ptid, 6 x variables, on y and one w

names(steam)

## [1] "ptid" "Y"    "x1"   "x2"   "x3"   "x4"   "x5"   "x6"   "W"   

colSums(is.na(steam))
## to check whether they have na

```

Now let us look into the correlations, the distributions in these variables.

```{r chunk2}
binary_vars <- names(steam)[
  sapply(steam, function(x) length(unique(x)) == 2)
]

continuous_vars <- names(steam)[
  sapply(steam, function(x) length(unique(x)) > 2)
]

binary_vars

## The binary variables are "Y"  "x5" "W" 

continuous_vars
## continuous variables are : "ptid" "x1"   "x2"   "x3"   "x4"   "x6" 

for (i in continuous_vars) {
  hist(steam[[i]],
       main=i,
       xlab=i,
       col="lightblue",
       border="white")
  
  
}

for (i in binary_vars) {
  cat("\n====================\n")
  cat(i,"\n")
  print(table(steam[[i]]))
  print(prop.table(table(steam[[i]])))
}
```
Now we have looked into the distributions ans prop tables of the variables.
Let us create the models.

```{r model}

model1<- glm(Y~ W +x1+x2+x3+x4+x5+x6,data = steam)

summary(model1)

## we got the AIC AIC: 146.6

confint(model1)

## Residuals

deviance_res <- residuals(model1, type = "deviance")

summary(deviance_res)

qqnorm(deviance_res)

plot(fitted(model1),
     deviance_res,
     xlab = "Predicted Probability",
     ylab = "Deviance Residuals",
     main = "Residuals vs Fitted")



## Now let us look into the CI for the odds ratio
CI <- exp(confint(model1))
OR <- exp(coef(model1))
pvals <- summary(model1)$coefficients[,4]


results <- data.frame(
  Variable = rownames(summary(model1)$coefficients),
  Coefficient = coef(model1),
  Odds_Ratio = exp(coef(model1)),
  CI_Lower = exp(confint.default(model1)[,1]),
  CI_Upper = exp(confint.default(model1)[,2]),
  P_value = summary(model1)$coefficients[,4]
)

results

```
Now let us create the discrimination and calibration for these plots

```{r model measures}

library(pROC)

pred_prob <- predict(model1, type = "response")

roc <- roc(steam$Y,pred_prob)

plot(roc,
     col = "blue",
     lwd = 3,
     main = "ROC Curve - Model 1")

auc(roc) ## 0.8042

```


Now let us create a model 2

```{r model2}

library(glmnet)
x <-model.matrix(
  Y ~ W + x1 + x2 + x3 + x4 + x5 + x6,
  data = steam
)[,-1] ## we removed -1 as glm automatically adds this

y <- steam$Y

model2 <- glmnet(
  x,
  y,
  family = "binomial",
  alpha = 1
)

## ploting 
plot(model2,
     xvar = "lambda",
     label = TRUE)

## Optimal validation

set.seed(123)

cv_lasso <- cv.glmnet(
  x,
  y,
  family = "binomial",
  alpha = 1,
  nfolds = 10
)

plot(cv_lasso)

# Predictions
lasso_prob <- predict(
  cv_lasso,
  newx = x,
  s = "lambda.min",
  type = "response"
)

# ROC
roc_lasso <- roc(y, as.vector(lasso_prob))

plot(roc_lasso,
     col = "red",
     lwd = 3)

auc(roc_lasso)

```
Sequential Parametric Synthetic Data Generation

```{r synthetic 1}
## Let us make a synthetic data using the parametric synthetic Data genertaion 

# Stage 1: Parametric Sequential Synthetic Data Generation


set.seed(123)

# Number of observations

n <- nrow(steam)

# Empty synthetic dataset

syn <- data.frame(
  ptid = 1:n
)

# STEP 1: Generate x1

mean_x1 <- mean(steam$x1, na.rm = TRUE)
sd_x1   <- sd(steam$x1, na.rm = TRUE)

syn$x1 <- rnorm(
  n,
  mean = mean_x1,
  sd = sd_x1
)

# STEP 2: Generate x2 | x1

model_x2 <- lm(x2 ~ x1, data = steam)

pred_x2 <- predict(model_x2, newdata = syn)

sigma_x2 <- summary(model_x2)$sigma

syn$x2 <- pred_x2 +
  rnorm(n, mean = 0, sd = sigma_x2)

# STEP 3: Generate x3 | x1 + x2

model_x3 <- lm(x3 ~ x1 + x2, data = steam)

pred_x3 <- predict(model_x3, newdata = syn)

sigma_x3 <- summary(model_x3)$sigma

syn$x3 <- pred_x3 +
  rnorm(n, mean = 0, sd = sigma_x3)

# STEP 4: Generate x4 | x1 + x2 + x3

model_x4 <- lm(x4 ~ x1 + x2 + x3, data = steam)

pred_x4 <- predict(model_x4, newdata = syn)

sigma_x4 <- summary(model_x4)$sigma

syn$x4 <- pred_x4 +
  rnorm(n, mean = 0, sd = sigma_x4)

# STEP 5: Generate x5 | x1 + x2 + x3 + x4

model_x5 <- glm(
  x5 ~ x1 + x2 + x3 + x4,
  family = binomial,
  data = steam
)

prob_x5 <- predict(
  model_x5,
  newdata = syn,
  type = "response"
)

syn$x5 <- rbinom(
  n,
  size = 1,
  prob = prob_x5
)

# STEP 6: Generate x6 | x1 + x2 + x3 + x4 + x5

model_x6 <- lm(
  x6 ~ x1 + x2 + x3 + x4 + x5,
  data = steam
)

pred_x6 <- predict(
  model_x6,
  newdata = syn
)

sigma_x6 <- summary(model_x6)$sigma

syn$x6 <- pred_x6 +
  rnorm(n, mean = 0, sd = sigma_x6)

# STEP 7: Generate Treatment W

model_W <- glm(
  W ~ x1 + x2 + x3 + x4 + x5 + x6,
  family = binomial,
  data = steam
)

prob_W <- predict(
  model_W,
  newdata = syn,
  type = "response"
)

syn$W <- rbinom(
  n,
  size = 1,
  prob = prob_W
)

# STEP 8: Generate Outcome Y

model_Y <- glm(
  Y ~ W + x1 + x2 + x3 + x4 + x5 + x6,
  family = binomial,
  data = steam
)

prob_Y <- predict(
  model_Y,
  newdata = syn,
  type = "response"
)

syn$Y <- rbinom(
  n,
  size = 1,
  prob = prob_Y
)


## synthetic dataset

str(syn)

names(syn)
```
Does the synthetic dataset (syn) preserve the important characteristics of the original dataset (d_steam)?

```{r check}
## dimensions

dim(steam)
dim(syn)

## Both have same , 150 observations for 9 variables 

## compare the mean and std deviations

real_summary <- data.frame(
  Variable = c("x1","x2","x3","x4","x6"),
  Mean = sapply(steam[,c("x1","x2","x3","x4","x6")], mean),
  SD = sapply(steam[,c("x1","x2","x3","x4","x6")], sd)
)

syn_summary <- data.frame(
  Variable = c("x1","x2","x3","x4","x6"),
  Mean = sapply(syn[,c("x1","x2","x3","x4","x6")], mean),
  SD = sapply(syn[,c("x1","x2","x3","x4","x6")], sd)
)

real_summary

syn_summary

## Now let us chacke the binary variables

prop.table(table(steam$x5))
prop.table(table(syn$x5))

prop.table(table(steam$W))
prop.table(table(syn$W))

prop.table(table(steam$Y))
prop.table(table(syn$Y))

## Npw let us compare the corelations

real_cor <- cor(
  steam[,c("x1","x2","x3","x4","x6")]
)

syn_cor <- cor(
  syn[,c("x1","x2","x3","x4","x6")]
)

round(real_cor,2)

round(syn_cor,2)
```
```{r check2}
par(mfrow=c(2,2))

hist(steam$x1,
     main="Real x1")

hist(syn$x1,
     main="Synthetic x1")

hist(steam$x2,
     main="Real x2")

hist(syn$x2,
     main="Synthetic x2")

par(mfrow=c(1,1))

```
```{r check3}
plot(
  density(steam$x1),
  lwd=2,
  main="Density of x1"
)

lines(
  density(syn$x1),
  col="red",
  lwd=2
)

legend(
  "topright",
  legend=c("Real","Synthetic"),
  col=c("black","red"),
  lwd=2
)
```
```{r check3}

real_model <- glm(
Y~W+x1+x2+x3+x4+x5+x6,
family=binomial,
data=steam
)

syn_model <- glm(
Y~W+x1+x2+x3+x4+x5+x6,
family=binomial,
data=syn
)

real_pred <- predict(
real_model,
type="response"
)

syn_pred <- predict(
syn_model,
type="response"
)

prop.table(table(real_pred,syn_pred))

```

Now let us compare discrimination

```{r check3}
library(pROC)

real_auc <- roc(
steam$Y,
real_pred
)

syn_auc <- roc(
syn$Y,
syn_pred
)

auc(real_auc)

auc(syn_auc)


## Now let us overlay the curves

plot(
real_auc,
col="blue",
lwd=2,
main="ROC Curves"
)

plot(
syn_auc,
add=TRUE,
col="red",
lwd=2
)

legend(
"bottomright",
legend=c("Real","Synthetic"),
col=c("blue","red"),
lwd=2
)
```
Now let us create a synthetic model with CART

```{r CART}
library(synthpop)

syn_data<- syn(steam, m=1, visit.sequence = c("x1","x2","x3","x4","x5","x6"), method = "cart", "cart", "cart", "cart","")

syn_data$predictor.matrix

## some features we found in the package

## m=0, gives , the predict.mtrix

##  predict.matric[W,x1] <- 0, gives the x1 removed form the prediction of variable W


## Now let us compare both the models.

head(syn_data)


```
```{r checkmodel2}

syn_cart <- syn_data$syn

real_model <- glm(
Y~W+x1+x2+x3+x4+x5+x6,
family=binomial,
data=steam
)

syn_model2 <- glm(
Y~W+x1+x2+x3+x4+x5+x6,
family=binomial,
data=syn_cart
)

real_pred <- predict(
real_model,
type="response"
)

syn_pred2 <- predict(
syn_model2,
type="response"
)

prop.table(table(real_pred,syn_pred))

```

Now let us compare discrimination

```{r check3}
library(pROC)

real_auc <- roc(
steam$Y,
real_pred
)

syn_auc2 <- roc(
syn_cart$Y,
syn_pred2
)

auc(real_auc)

auc(syn_auc2)


## Now let us overlay the curves

plot(
real_auc,
col="blue",
lwd=2,
main="ROC Curves"
)

plot(
syn_auc2,
add=TRUE,
col="red",
lwd=2
)

legend(
"bottomright",
legend=c("Real","Synthetic"),
col=c("blue","red"),
lwd=2
)
```

Now we will do a sequential comparison

```{r checkseq}
set.seed(17914709)

syn_data <- syn(
  steam,
  method = "cart",
  m = 5,
  visit.sequence = c(
    "x1","x2","x3","x4","x5","x6","W","Y"
  )
)

compare(syn_data, steam)
```

Now let us think about  