---
title: "syn_2"
author: "Finta_Francis_R_project"
date: "2026-07-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(synthpop)
library(readr)
library(pROC)
library(partykit)

## Here we are trying to explore CTree

## First we will load the dataset GlaucomaM
head(GlaucomaM)

str(GlaucomaM)

table(GlaucomaM$Class)
```
Now let us create a Ctree

```{r chunk1}
gtree <- ctree(Class ~ .,
  data = GlaucomaM)

gtree
```

## Now et us dive deep into the dataset

```{r chunk2}
plot(gtree)

info_node(node_party(gtree))

split_node(node_party(gtree))

## Let us visualize the split

plot(
  GlaucomaM$vari,
  as.numeric(GlaucomaM$Class),
  xlab="vari",
  ylab="Class"
)

abline(v = 0.059)
```
Now let us contro alpha a bit more 

```{r chunk3}
gtree2 <- ctree(
  Class ~ .,
  data = GlaucomaM,
  control = ctree_control(alpha = 0.001)
)

plot(gtree2)

```

Let us understand the alogorithm better

```{r chunk4}
set.seed(123)

sim <- data.frame(
  age = rnorm(200, 50, 10),
  biomarker = rnorm(200),
  treatment = rbinom(200, 1, 0.5)
)

sim$outcome <- ifelse(
  sim$biomarker + sim$treatment > 0,
  "Disease",
  "Normal"
)

sim$outcome <- factor(sim$outcome)

head(sim)

table(sim$outcome)

tree_default <- ctree(
  outcome ~ .,
  data = sim
)

tree_default
plot(tree_default)

nodeids(tree_default)

## Let us check the child nodes
info_node(
  node_party(tree_default), id = 2
)

## predict the nodes
predict(
  tree_default
)

predict(
  tree_default,
  type="prob"
)

predict(
  tree_default,
  type="node"
)

## Matrix

table(
  predict(tree_default),
  sim$outcome
)

## misclassification rate
mean(
  predict(tree_default) != sim$outcome
)
```


## Now let us understand the test control

```{r chunk5}

tree_max <- ctree(
  outcome ~ .,
  data = sim,
  control = ctree_control(
    teststat = "max"
  )
)

tree_max

plot(tree_max)
```

## Now let us use :
##. Uses: c quad =(t−μ)Σ −1(t−μ) T

```{r chunk6}
tree_quad <- ctree(
  outcome ~ .,
  data = sim,
  control = ctree_control(
    teststat = "quad"
  )
)

tree_quad

plot(tree_quad)
```
```{r chunk7}
tree_bonf <- ctree(
  outcome ~ .,
  data = sim,
  control = ctree_control(
    testtype = "Bonferroni"
  )
)

plot(tree_bonf)
```
Now let us look into random forest genration

```{r random forest}

data(iris)

data_iris <- iris

set.seed(213456)
library(synthpop)

set.seed(213456)

synth_cart <- syn(
  data_iris,
  method = "cart",
  m = 1
)

cart_data <- synth_cart$syn

head(cart_data)
```


## Now let us create the same dataset using random forest

```{r random forest1}
set.seed(213456)

synth_rf <- syn(
  data_iris,
  method = "rf",
  m = 1
)

rf_data <- synth_rf$syn

head(rf_data)
```

# Now let us compare the statistical utlitiy in this models

```{r stats}
## First we will look into the summary 

summary(cart_data)
summary(rf_data)
## This seems almost identical to the original dataset

summary(data_iris)

## Now let us compare the means 

library(dplyr)

compare_means <- data.frame(
 Variable = c(
 "Sepal.Length",
 "Sepal.Width",
 "Petal.Length",
 "Petal.Width"
 ),

Original = sapply(
 data_iris[,1:4],
 mean
 ),

CART = sapply(
 cart_data[,1:4],
 mean
 ),

RF = sapply(
 rf_data[,1:4],
 mean
 )
)

compare_means
```
## ## Now we will compare the SD

```{r stats_SD}
library(dplyr)
compare_sd <- data.frame(
 Variable = c(
 "Sepal.Length",
 "Sepal.Width",
 "Petal.Length",
 "Petal.Width"
 ),

Original = sapply(
 data_iris[,1:4],
 sd
 ),

CART = sapply(
 cart_data[,1:4],
 sd
 ),

RF = sapply(
 rf_data[,1:4],
 sd
 )
)

compare_sd
```

##  Now let us compare them graphically

```{r graphs}
library(tidyverse)

original <- data_iris %>%
  mutate(Source = "Original")

cart <- cart_data %>%
  mutate(Source = "CART")

rf <- rf_data %>%
  mutate(Source = "Random Forest")


combined <- bind_rows(
  original,
  cart,
  rf
)


## Then plot it using ggplot
library(ggplot2)

combined_long <- combined %>%
  pivot_longer(
    cols = 1:4,
    names_to = "Variable",
    values_to = "Value"
  )


ggplot(
  combined_long,
  aes(
    x = Value,
    fill = Source
  )
) +
  geom_density(alpha = 0.4) +
  facet_wrap(
    ~Variable,
    scales = "free"
  ) +
  theme_bw() +
  labs(
    title = "Distribution comparison",
    x = NULL,
    y = "Density"
  )
```

## Let us create some box plots

```{r graphs2}
ggplot(
  combined_long,
  aes(
    x = Source,
    y = Value
  )
) +
  geom_boxplot() +
  facet_wrap(
    ~Variable,
    scales = "free"
  ) +
  theme_bw() +
  labs(
    title = "Boxplot comparison of synthetic and original data"
  )
```
## Mean preservation plot

```{r graphs3}
mean_comparison <- combined_long %>%
  group_by(Source, Variable) %>%
  summarise(
    Mean = mean(Value),
    .groups="drop"
  )


ggplot(
  mean_comparison,
  aes(
    x = Variable,
    y = Mean,
    group = Source,
    color = Source
  )
) +
  geom_point(size=3) +
  geom_line() +
  theme_bw() +
  labs(
    title="Mean preservation across datasets"
  )
```
PCA plot
```{r graphs4}
library(ggfortify)

pca_input <- combined %>%
  dplyr::select(
    Sepal.Length,
    Sepal.Width,
    Petal.Length,
    Petal.Width
  )


pca_result <- prcomp(
  pca_input,
  scale.=TRUE
)


autoplot(
  pca_result,
  data=combined,
  colour="Source"
) +
theme_bw() +
labs(
 title="PCA comparison of original and synthetic datasets"
)
```
```{r stats4}
cor_real <- cor(data_iris[,1:4])

cor_cart <- cor(cart_data[,1:4])

cor_rf <- cor(rf_data[,1:4])


difference_cart <- abs(
  cor_real - cor_cart
)

difference_rf <- abs(
  cor_real - cor_rf
)

ks.test(
 data_iris$Sepal.Length,
 cart_data$Sepal.Length
)

ks.test(
 data_iris$Sepal.Length,
 rf_data$Sepal.Length
)


```

